Concurrent Programming with CXL

### 2020-06-04 book/paper.tex
Latest

But model checking can also help with proving an algorithm correct.
The reason is that the correctness of many programs is built around
\emph{invariants}: predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
those invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof.

@@

Like Python, CXL is an imperative,
dynamically typed and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Every statement in CXL must be terminated by a semicolon
(even \texttt{while} statements and \texttt{def} statements);
in Python semicolons are optional and rarely used;
\item Correct indentation in CXL is encouraged but optional;
\item Python is object-oriented; CXL is not;
\item CXL does not (currently) support floating point;
\item CXL does not support I/O of any kind;
\item CXL does not support operator precedence or associativity.  Use
parentheses liberally to remove ambiguity.
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in CXL.

@@

``Running'' this CXL program will try all possible executions, which
includes all possible values for $x$.  Essentially, the $\texttt{choose}(S)$
operator provides the input to the program by selecting some value from the
set~$S$, while the $\texttt{assert}$ statement checks that the output is
correct.  If the program is correct, the output of CXL is the size of the
``state graph'' and its diameter (more on this later).  If not, CXL also
reports what went wrong, typically by displaying a summary of an execution in
which something went wrong.

\begin{quote}
Try it out!
\begin{itemize}
\item See what happens if, instead of initializing \texttt{result} to 0,
you initialize it to 1.
\item Write a similar program that computes squares or factorials.
\end{itemize}
\end{quote}

@@

\begin{figure}
\begin{code}
\begin{verbatim}
Up.cxl:1 def incrementer(self):
   0 Jump 39
   1 Frame ('incrementer', 'Up.cxl', 1, 5) [('self', 'Up.cxl', 1, 17)] 12
Up.cxl:2     count = count + 1;
   2 Constant (1, 'Up.cxl', 2, 21)
   3 Name ('count', 'Up.cxl', 2, 13)
   4 Load 1
   5 Nary ('+', 'Up.cxl', 2, 19) 2
   6 Name ('count', 'Up.cxl', 2, 5)
   7 Store 1
Up.cxl:3     done[self] = True;
   8 Constant (True, 'Up.cxl', 3, 18)
   9 LoadVar ('self', 'Up.cxl', 3, 10)
   10 Name ('done', 'Up.cxl', 3, 5)
   11 Store 2
   12 Return
\end{verbatim}
\end{code}
\caption{The first part of the machine code corresponding to Figure~\ref{fig:inc}.}
\label{fig:inccode}
\end{figure}

\begin{figure}
\begin{code}
\begin{verbatim}
==== Safety violation ====
incrementer/0 [1-6]          7 dict{ .count:0, .done:dict{ 0:False, 1:False } }
incrementer/1 [1-11]        12 dict{ .count:1, .done:dict{ 0:False, 1:True  } }
incrementer/0 [7-11]        12 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
main/() [14-19,22-27,29-36] 37 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
\end{verbatim}
\end{code}
\caption{The output of running Figure~\ref{fig:inc}.}
\label{fig:incoutput}
\end{figure}

@@

The \texttt{Frame} instruction has three arguments:
\begin{enumerate}
\item A token containing the name of the method;
\item A list of tokens containing the names of the arguments of the method;
\item A program counter of the method's \texttt{Return} instruction.
\end{enumerate}

A \emph{token} is a four-tuple containing a \emph{lexeme}, a file name, a
line number in the file, and a column number in the file.  For example,
\texttt{(’up’, ’Up.cxl’, 1, 5)} represents the lexeme \texttt{up} in
file \texttt{Up.xcl}, line~1, column~5.  Tokens make it easy to relate
the CXL machine instructions to the CXL source code.

@@

\begin{enumerate} \setcounter{enumi}{1}
\item the \texttt{Constant} instruction pushes the constant 1
onto the stack of the process.
\item The \texttt{Load} instruction loads the shared variable
\texttt{count} from the shared memory and pushes its value onto the
stack.
\item \texttt{Nary} stands for \texttt{n-ary}: in this case it is a \texttt{+}
operation with 2 arguments.  It pops two values from the stack, adds them,
and pushes the result back onto the stack.
\item The \texttt{Name} instruction pushes the address of the \texttt{count}
variable onto the stack.
\item The \texttt{Store} instruction pops the address of a variable and pops
a CXL value, and updates the memory by replacing the CXL value oat the given
address with the new CXL value.
\end{enumerate}

@@

If we look at the middle three rows, we see that:
\begin{enumerate}
\item Process 0 executed instructions 1 through 5, loading the value of
\texttt{count} but stopping just before storing 1 into \texttt{count};
\item Process 1 executed instructions 1 through 10, storing 1 into
into \texttt{count} and storing \texttt{True} into \texttt{done[2]};
\item Process 0 continues execution, storing value 1 into \texttt{count}
and storing \texttt{True} into \texttt{done[1]}.
\end{enumerate}

>>>>

But model checking can also help with proving an algorithm correct.
The reason is that the correctness of many programs is built around
\emph{invariants}: predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
those invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof.
Throughout this book we will include, as much as possible, examples
of such invariants as they often provide excellent insight into
why a particular algorithm works.

@@

Like Python, CXL is an imperative,
dynamically typed, and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Every statement in CXL must be terminated by a semicolon
(even \texttt{while} statements and \texttt{def} statements);
in Python semicolons are optional and rarely used;
\item Correct indentation in CXL is encouraged but optional;
\item Python is object-oriented; CXL is not;
\item CXL does not (currently) support floating point;
\item CXL does not support I/O of any kind;
\item CXL only supports basic operator precedence or associativity.
Use parentheses liberally to remove ambiguity.
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in CXL.

@@

``Running'' this CXL program will try all possible executions, which
includes all possible values for $x$.  Try it out:

\begin{code}
\begin{verbatim}
$ ./cxl triangle.cxl
#states = 13
no issues found
$
\end{verbatim}
\end{code}

Essentially, the $\texttt{choose}(S)$
operator provides the input to the program by selecting some value from the
set~$S$, while the $\texttt{assert}$ statement checks that the output is
correct.  If the program is correct, the output of CXL is the size of the
``state graph'' (13 states in this case).  If not, CXL also
reports what went wrong, typically by displaying a summary of an execution in
which something went wrong.

\begin{quote}
Experiment!
\begin{itemize}
\item See what happens if, instead of initializing \texttt{result} to 0,
you initialize it to 1.
\item Write a similar program that computes squares or factorials.
\end{itemize}
\end{quote}

@@

\begin{figure}
\begin{code}
\begin{verbatim}
Up.cxl:1 def incrementer(self):
    0 Jump 39
    1 Frame incrementer ['self'] 12
Up.cxl:2     count = count + 1;
    2 Push 1
    3 PushAddress count
    4 Load 1
    5 2-ary +
    6 PushAddress count
    7 Store 1
Up.cxl:3     done[self] = True;
    8 Push True
    9 PushVar self
   10 PushAddress done
   11 Store 2
   12 Return
\end{verbatim}
\end{code}
\caption{The first part of the machine code corresponding to Figure~\ref{fig:inc}.}
\label{fig:inccode}
\end{figure}

\begin{figure}
\begin{code}
\begin{verbatim}
CXL Assertion failed ('assert', 'Up.cxl', 9, 5) 1
#states = 79
==== Safety violation ====
__init__/() [0,39-56]       57 dict{ .count:0, .done:dict{ 0:False, 1:False } }
incrementer/0 [1-6]          7 dict{ .count:0, .done:dict{ 0:False, 1:False } }
incrementer/1 [1-11]        12 dict{ .count:1, .done:dict{ 0:False, 1:True  } }
incrementer/0 [7-11]        12 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
main/() [14-19,22-27,29-36] 37 dict{ .count:1, .done:dict{ 0:True,  1:True  } }
\end{verbatim}
\end{code}
\caption{The output of running Figure~\ref{fig:inc}.}
\label{fig:incoutput}
\end{figure}

@@

The \texttt{Frame} instruction has three arguments:
\begin{enumerate}
\item The name of the method;
\item A list containing the names of the arguments of the method;
\item The program counter of the method's \texttt{Return} instruction.
\end{enumerate}

@@

\begin{enumerate} \setcounter{enumi}{1}
\item the \texttt{Push} instruction pushes the constant 1
onto the stack of the process.
\item the \texttt{PushAddress} instruction pushes the address of
the shared variable \texttt{count} onto the stack.
\item The \texttt{Load} instruction pops the address of the
\texttt{count} variable and then pushes the value of the
\texttt{count} variable onto the stack.
\item \texttt{2-ary} is a \texttt{+} operation with 2 arguments.
It pops two values from the stack (1 and the value of \texttt{count}),
adds them, and pushes the result back onto the stack.
\item The \texttt{PushAddress} instruction pushes the address of the \texttt{count}
variable onto the stack.
\item The \texttt{Store} instruction pops the address of a variable and pops
a CXL value (the sum of the \texttt{count} variable and 1), and updates the
\texttt{count} variable.
\end{enumerate}

@@

If we look at the middle three rows, we see that:
\begin{enumerate}
\item Process 0 executed instructions 1 through 7, loading the value of
\texttt{count} but stopping just before storing 1 into \texttt{count};
\item Process 1 executed instructions 1 through 12, storing 1 into
into \texttt{count} and storing \texttt{True} into \texttt{done[2]};
\item Process 0 continues execution, storing value 1 into \texttt{count}
and storing \texttt{True} into \texttt{done[1]}.
\end{enumerate}
